\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-01 paper.tex

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
An ARG is classically defined as a graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of a sample of genomes,
encoding an inheritance structure where
crossover breakpoints are associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of recent ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show that the gARG encoding can fully capture the richness of large-scale datasets
and is compatible with modern simulation and inference methods.
We also show that, unlike the eARG encoding which requires
% TODO rephrase "to be detailed" is ambiguous
all recombination events to be encoded,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species or higher grouping. 
This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true ARG unquestionably exists,
it can also never be entirely known.
% TODO: I'm not happy with the phrasing here, anyone want to jump in?
Even if the mutation rate were high enough to identify every
recent branching point uniquely (while remaining biologically survivable),
such a high rate would saturate the genome with mutations, inevitably obscuring
past history.
% Must be a citation for this?
Similarly, it is unlikely that the details
of ancient recombinations can ever be known with much precision.

% Popgen models and the species ARG
Species-level ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}, and
agricultural datasets are even
richer in terms of generations~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
While we are not sequencing entire species,
we are certainly sampling at a scale where the assumptions of
population genetics models are substantially broken.
By necessity and design, such mathematical models are caricatures
of biological reality,
allowing us to draw general conclusions about population processes
without the burden of unnecessary detail.
 % TODO we're singling out the coalescent here, but it's just an 
 % exmaple. The DTWF is no better an approximation of the last 5
 % generations of UKB, e.g.
The coalescent with recombination (henceforth, ``coalescent'')
is a powerful model that defines a distribution over ARGs
for a given set of parameters.
A key assumption of the coalescent is that
sample size is much less than the effective population size ($N_e$)
and although the model has proved surprisingly robust to violations
of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
nelson2020accounting},
sample size can now be an order of magnitude or more \emph{larger} than the
usually assumed $N_e$ values.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the full complexities
of family structure or the vagaries of history inherent in these
datasets.

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
% TODO rephrase heavily!
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

We address these problems here by providing a simple formal description of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% Do we need to say this? Probably not?
We are not questioning the form of the true ARG, or its existence.

>>>>

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
% TODO rephrase, this is wooly now
An ARG is classically defined as a graph of
coalescence and recombination events in the history of a sample of genomes,
% TODO this is vague
encoding an inheritance structure where
crossover breakpoints are associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of recent ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show that the gARG encoding can fully capture the richness of large-scale datasets
and is compatible with modern simulation and inference methods.
% TODO rewrite
We also show that, unlike the eARG encoding which requires
all recombination events to be encoded,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
% Note: I started off talking about the "ultimate" ARG of everone
% who ever lived, but didn't find it that helpful.
True ARGs are the actual history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although population ARGs unquestionably exist,
they can also never be entirely known because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

% Population ARGs are here, now.
Population ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such population-scale sequencing
projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}.
Agricultural datasets are on a similar scale,
and also include dense multi-generational sampling
and near-perfect pedigree
information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
Recent advances have made it possible to actually
\emph{estimate} ARGs at population scale:
ARGs have already been inferred for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
While this new population-scale reality presents many
exciting opportunities, it also poses substantial challenges
to existing methodologies.

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally, these formalisms assume that an
entire ARG is known with complete precision,
and are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be known in complete detail,
there is currently a substantial gap between our theoretical frameworks
and practical application.
The breakthroughs in scale achieved by recent
% Tedious to cite again, but I don't want to leave Relate out of the
% motivatory discussion
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based (in different ways) on inferring
approximate \emph{structures},
instead of a complete, fully detailed history.
% Not sure the point is coming across that well here, but it's a
% really important point and a significant
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent. % with recomb is surely unnecessary here
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description
that is sufficiently general.

We address these problems here by providing a simple formal description of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% True ARGs, being perfectly resolved, can ultimately be described in
% either formalism
